---
title: "Histograms"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Histograms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.width = 8,
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ggpower)
library(scales)  # for trans_new
library(tidyverse)
library(cowplot)
```


Histograms and axis-transformations are a non-trivial combination,
perhaps because people are not aware what histograms are, or because
of the subtle difference between ggplot2's *scale_x_log10* and *coord_trans*.

# Bar height != number of observations

I used to think so too, but the **height** of each histogram bar does NOT indicate the number
of observations in that bin. It is the bar's **area** that indicates that number.
This has a beautiful consequence: the bar height is not an absolute number, but
an estimate of density in the bin - histograms thus show estimates of the 
probability density irrespective of binwidth.


# Why coord_trans is better than scale_x_continuous


**tl;dr** (too long, didn't read):

  * *scale_x_continuous(trans="log")* does not make histogram and probability
    density function (pdf) fit together, *coord_trans(trans="log")* does this
  * *coord_trans* requires breaks that start above zero (to avoid errors) and
  are exponentially spaced, not linearly (makes bars equal widths)


It is not uncommon to have the histogram of some variable x and the (allegedly) underlying
theoretical distribution in one plot. On the linear scale, this is done easily
by combining *geom_histogram* and *geom_line*. Note how the code creating
*xseq* seems a bit complicated and for geom_line we could use a linear sequence
as well. For geom_histogram, however, using exponential, non-zero
breaks in the histogram is required for using *coord_trans(trans="log")*, 
so we do it already here.

```{r}
# generate data from two log-normal distributions
dat <- data.frame(class=c("a","b"),
                  obs = rlnorm(1000, meanlog = c(0, 2), sdlog = c(.5, .5)))
# xseq: exponentially spaced and avoiding zero:
xseq <- pmax(1e-5, exp(seq(log(.9*min(dat$obs)), log(1.1*max(dat$obs)), length.out = 100)))
# empirical histogram and theoretical density in same plot:
p <- ggplot() +
  geom_histogram(data=dat, aes(obs, stat(density)), breaks = xseq)+
  geom_line(aes(xseq, .5*dlnorm(xseq, 0, .5))) + # half of dat is from this distribution
  geom_line(aes(xseq, .5*dlnorm(xseq, 2, .5)))   #  other half from herep
p
```

We see two log-normal distributions, and it would make sense to display them
on the log-scale:
```{r}
plot_grid(p + ggtitle("linear"),
          p+scale_x_continuous(trans = "log")+ggtitle("scale_x_continuous"),
          p+coord_trans(x="log")+ggtitle("coord_trans"), ncol=3 )
```

Evidently, log-transforming with scale_x_log10 leads to discrepancies between
the histogram (empirical density) and the pdf (theoretical density). 
The help of *coord_trans* tells us in the Examples the reason:
scale_x_* transform BEFORE statistics and coord_trans after. In our example,
this means that scale_x creates a histogram of log(x), which has a different
density function than x. Coord_trans, on the other hand, does what we want because
it first computes the histogram and than transforms the coordinate system.

## Note on exponentially spaced, non-zero *breaks*

We created *xseq* above somewhat complicated so that coord_trans works and 
here is the reason.

Log-transforming histograms is tricky because the left boundary of the
left-most bin can be a negative number, which causes errors with *coord_trans*
(while scale_x_continuous is fine).
One could use *trans_new* to modify *log_trans* (e.g. replacing 0s with a small 
non-zero number), but I found it more useful
to use *geom_histogram*'s **breaks** argument to define the bin borders, starting
from a small positive and non-zero number. While we are at it, we can make those
breaks exponentially-spaced, as linear spacing leads to very broad left bars,
masking information. I do this above with *xseq*, which is then passed to 
geom_histogram as *breaks*.


# Using ggpower makes it easy

Let's do the plot from above again, but this time with nicer breaks.
We will also sqrt-transform the y-axis to make differences in densities
less pronounced:
```{r}
p+coord_trans(x="log", y = "sqrt") + scale_x_continuous(breaks = function(lims) power_breaks(lims, .5, 100), labels = semi_scientific_formatting)
```


