---
title: "lognormal_distribution"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{lognormal_distribution}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ggplot2)
library(ggpower)
library(scales)
# library(tidyverse)
library(cowplot)
```

The log-normal distribution plays an important role in biology
(and everywhere else: finances, engineering, etc.) due
to the Central Limith Theorem: if many random effects are added together, the 
result converges to a normal distribution, if they are **multiplied** it converges
to a log-normal distribution.

# The meanlog argument
Docu says that *meanlog* and *sdlog* are the mean and standard deviation
of the distribution on the log scale, and here we look at what that means.




## The log of a normal
If x is log-normally distributed, log(x) is normal with mean *meanlog* and 
variance $sdlog^2$:
```{r}
x <- rlnorm(100000, meanlog = 2, sdlog = 1.5)
hist(log(x), 100, freq = F)
points(seq(-4,8, length.out = 50), dnorm(seq(-4, 8, length.out = 50), mean = 2, sd = 1.5),
       type="l")
```
This is obvious from the Lognormal's definition and helpful if log-transforming
your data is an option.
Let's look at building intuition for the *meanlog* argument in Lognormal
distribution itself, i.e. the untransformed data.


## meanlog is the logged median value

I compute mean, median and mode for given meanlog (mu1) and sdlog (s1) and plot
it into the density function.

```{r}
mu1 = 2; s1  = 0.5
lnorm_stats <- c(mode   = exp(mu1 - s1*s1),  # see e.g. wikipedia
                 median = exp(mu1),
                 mean   = exp(mu1+s1*s1/2))
# logarithmic spaced points on x-axis are not crucial here, they become useful
# when we do histograms, see below
xseq <- pmax(1e-5, exp(seq(log(1), log(25), length.out = 1000)))
p <- ggplot() +
  geom_line(aes(xseq, dlnorm(xseq, mu1, s1))) +
  geom_vline(aes(xintercept = lnorm_stats,
                 color = names(lnorm_stats)), linetype = "dashed") +
  xlab("x") + ggtitle(paste0("dlnorm(x, meanlog=", mu1, ", sdlog=", s1, ")")) +
  scale_color_hue(name = "lognorm stats",
                  labels = c("mean(x) = exp(meanlog+sdlog^2/2)",
                             "median(x) = exp(meanlog)",
                             "mode(x) = exp(meanlog-sdlog^2)")) +
  theme(legend.justification = c(1, 0), legend.position = c(.9, .5))

p
```

We see that *meanlog* has the most direct relationship with the expected
median of a Lognormally distributed random variable X.
Put differently, *meanlog* is the mean of log(X), **not the mean of X itself**.
Or simply: *meanlog* is X's logged median value.

I'm repeating this on purpose because confusion arises easily when we log-transform
the x-axis:
```{r}
p+scale_x_log10() + theme(legend.justification = c(1, 0), legend.position = c(.4, .5))
```

On the log-scale, x looks like a normal with mean $meanlog-sdlog^2$.
Note the difference to the above histogram of $log(x)$, which follows a 
normal distribution with mean $meanlog$.





## Summary


So the *meanlog* argument has two intuitions for a log-normal distributed x: 
  
  * *median*$(x) = e^{meanlog}$
  * *mean*$(log(x)) = meanlog$


```{r}
x <- rlnorm(100000, meanlog = 2, sdlog = 1.5)
c(
  mean_x_logged =log( mean(x) ), # NOT 2, but instead exp( meanlog + sdlog^2/2 ) = 3.125
  median_x = median(x),          # exp(2) = 7.38
 # once you log-transform x, things are easy
  mean_logx= mean(   log(x) ),  # meanlog, i.e. 2
  sd_logx  = sd( log(x) )       # sdlog, i.e. 1.5 
  )    


```




